---
title: "Smoothing II"
format: html
---


## Applied Mathematics in Biology: Single Cell Omics

Lecture Notes

Simon Anders, Univ. Heidelberg, 2023-07

## Smoothing II (Preliminary Notes)

### Standard preprocessing

The packages we need today:

```{r}
suppressPackageStartupMessages({
  library( Matrix )
  library( sparseMatrixStats )
  library( Seurat )
  library( tidyverse )
  library( locfit ) })
```

We again use the IFNAGRKO data set

```{r}
cells <- read.csv( "/mnt/raid/ifnagrko/ifnagrko_obs.csv", row.names = 1 )
genes <- read.csv( "/mnt/raid/ifnagrko/ifnagrko_var.csv", row.names = 1 )
counts <- readMM( "/mnt/raid/ifnagrko/ifnagrko_raw_counts.mtx.gz" )
rownames(counts) <- rownames(cells)
colnames(counts) <- genes$gene_name
counts <- t(counts)
```
First the standard preprocessing steps.

Log-normalization:

```{r}
fracs <- t( t(counts) / colSums(counts) )
lfracs <- log1p( fracs*1e4 )
```

Finding highly variable genes:

```{r}
rowVars( fracs, useNames=TRUE ) %>% sort(decreasing=TRUE) %>%
  head(1500) %>% names() -> hvg
```

PCA:

```{r}
pca <- irlba::prcomp_irlba( t(lfracs[hvg,]), 20, work=200 )
rownames(pca$x) <- colnames(lfracs)
rownames(pca$rotation) <- hvg
```

UMAP:

```{r}
set.seed( 12345678 )
nn <- FNN::get.knn( pca$x, 30 )
names(nn) <- c( "idx", "dist" )

ump <- uwot::umap( pca$x, nn_method=nn, min_dist=.2 )
plot( ump, pch=".", asp=1, col=1+(cells$celltype2=="qNSC1") )
```

In the UMAP plot, we have highlighted the quiescent neuronal stem cells, to
mark the beginning of the lineage.


### Recap: Smoothing by local regression

We make a table of expression data for aquaporin 4 (Aqp4, an astrocyte marker, present 
in the NSCs)

```{r}
tibble(
  cell = colnames(counts),
  count = counts[ "Aqp4", ],
  total = colSums(counts),
  celltype = cells$celltype,
  lineage = cells$celltype1 == "lineage",
  pt = cells$dpt_pseudotime,
  jitter = 10^runif( ncol(counts), -5, -4.7 ) 
) -> tbl0

head( tbl0 )
```

As before, we smooth along pseudotime. So, let's have a look at it again:

```{r}
cbind( 
  as_tibble(ump),
  tbl0 ) %>%
ggplot() + 
  geom_point(aes(x=V1,y=V2,col=pt), size=.1) +
  coord_equal() + scale_color_viridis_c()
```
The pseudotime is defined only within the lineage, so we subset to these cells:

```{r}
tbl0 %>% filter( lineage ) -> tbl
```

Let's plot the expression fractions for Aqp4 versus the pre-defined pseudotime:

```{r}
plot( count/total + 1e-5 ~ pt, tbl, log="y", cex=.1 )
```

To make the zeroes better visible, we replace them with the "jitter", i.e., the small 
random numbers that we put above in the column "jitter".

```{r}
plot( pmax( count/total, jitter ) ~ pt, tbl, log="y", cex=.1 )
```

Let's smooth with local regression (locfit):

We prepare a grid of 1000 values between 0 and 1 (ptg), and ask for smoothed 
fitted values for these by running locfit as follows:

```{r}
ptg <- seq( 0, 1, l=1000 )

fit <- locfit( 
  count ~ lp( pt, nn=.2 ), tbl, 
  family="poisson", base=log(tbl$total), ev=ptg )

fitted_curve = tibble( x=ptg, y=predict(fit) )
head( fitted_curve )
```

With this call, locfit does local regression, fitting quadratic polynomials
locally via GLM with Poisson likelihood and log link, using a kernel with 
adaptive bandwidth such that the kenrel always covers a fraction of 0.2 
of the cells. We have regressed the counts onto the pseudotime, using the log totals
as offsets.

Here's a plot of the fitted curve:
```{r}
plot( pmax( count/total, jitter/10 ) ~ pt, tbl, log="y", cex=.1, ylim=c(1e-6, 5e-3) )
lines( y ~ x, fitted_curve, col="magenta" )
```
### Deviance

We call locfit ones more, this time asking for evaluation not along our value grid `ptg`, but
for the cells' actual pseudotime values, and then write the fitted values into a new
table column `pred_expr`:

```{r}
fit <- locfit( count ~ lp( pt, nn=.2 ), tbl, family="poisson", base=log(tbl$total), ev=tbl$pt )
tbl$pred_expr <- predict(fit)

head(tbl)
```

Let's write $k_i$ for the UMI count (`count`)for our gene (Aqp4) for cell $i$, $s_i$ for the cell's total count (`total`)
and $\hat y_i$ for the fitted curve value (`pred_expr`). The log likelihood that was maximized by `locfit` is the expression
$\sum_i\log f_\text{Pois}(k;\hat y_i s_i)$, where $f_\text{Pois}$ is the Poisson probability mass function, i.e., the $f_\text{Pois}(k,\lambda)$
is the for a random variable that is Poisson distributed with expectation $\lambda$ to take the value $k$: $f_\text{Pois}(k,\lambda) = \lambda^k e^{-\lambda}/k!$. In R, the function `dpois` calculates this fucntion, or (if one sets log=TRUE) its logarithm.

It is useful to compare for each cell the log likelihood of a fitted value $\hat\lambda_i=s_i\hat y_i$, i.e., $\log f_\text{Pois}(k,\lambda_i)$ with the maximum likelihood that could be achieved if $\lambda_i$ could be chosen freely. The log likelihood function $\log f_\text{Pois}(k,\lambda)$ takes its
maximum w.r.t. to $\lambda$ for $\lambda=k$, and this is then called the likelihood of the saturated model. The difference between the attained maximum
likelihood and the saturated likelihood tells us something about the goodness of fit. Hence, we define
$$ D_i = -2\left( \log f_\text{Pois}(k_i,\lambda_i) - \log f_\text{Pois}(k_i,k_i) \right) $$
as the *deviance* of the fit for cell $i$.

Why is there a prefactor $-2$? TO see, consider how $D-I$ would look if we had done ordinary least square (OLS) regression, i.e.,
tried to maximize a normal likelihood in order to get a fitted value $\hat y_i$ clsoe to an observed value $y_i$. Then, we would
write $D_i = -2\left( \log f_\mathcal{N}(y_i;\hat y_i, \sigma) - \log f_\mathcal{N}(y_i; y_i, \sigma) \right)$, which, after inserting the 
normal density $f_\mathcal{N}(y;\mu, \sigma) = e^{-(y-\mu)^2/2\sigma^2}/\sqrt{2\pi\sigma^2}$, simplifies to $D_i = (y_i-\hat y_i)^2/\sigma^2$.
The prefactor 2 cancels against the 2 in front of $\sigma^2$ in the normal pdf, and we are left with the square of $z=(y_i-\hat y_i)/\sigma$, i.e.,
of the residual scaled by the standard deviation, which is often called the $z$-value. 

So, maximizing the likelihood in a GLM fit means minimizing the deviance, which is an generalization of the sum of squares that we minimize in
OLS regression. The least-square criterion is motivated by it minimizing the residual variance, and this explains why we call $D_i$ the "deviance",
a generalization of the residual variance.

The connection to the $z$ value motivated transforming the deviance into a "generalized z value" or "deviance residual" 
by taking the square root and putting a sign as needed to indicate whether the data point lies above or below the fitted curve:

$$ z_i = \sqrt{D_i}\, \operatorname{sgn}(k_i-s_i\hat y_i). $$
Here, we add deviance and deviance residual:

```{r}
tbl %>%
mutate( deviance = -2 * ( 
  dpois( count, total * pred_expr, log=TRUE ) - 
  dpois( count, count, log=TRUE ) ) ) %>%
mutate( devres = sqrt(deviance) * sign( count - total*pred_expr ) ) %>%
mutate( y = pmax( count/total, jitter/10 ) ) -> tbl

head( tbl )
```

Let's plot our fit again, now coloring the points by deviance residual:

```{r}
ggplot(tbl) + 
  geom_point( aes( x=pt, y=y, col=devres ), size=.3 ) +
  geom_line( aes( x=x, y=y ), data=fitted_curve, lty="dotted" ) +
  scale_y_log10() +
  scale_color_gradient2( mid="gray80", limits=c(-2.5,2.5), oob=scales::oob_squish ) +
  theme_bw()
```
Here's a histogram of the deviance residuals:

```{r}
hist( tbl$devres, 100 )
```
This now allows us to judge whether the remaining variance or deviance, that has not
been explained by the fit is mostly due to Poisson variation. If this were the case,
this histogram should look roughly standard-normal -- if not in shape then at least with respect to
fastness of the tails.

A QQ plot help judging this:

```{r}
qqnorm( tbl$devres )
abline( 0, 1 )
```

There clearly is variation in excess of what we expect due to the Poisson distribution

### Negative binomial instead of Poisson

Let's assume that the biological variation is 10%. By this we mean that given a sample
of cells with the same pseudotime values, we expect their pre-Poisson expression, i.e.,
their values $y_i$ that go into the Poisson, vary with a coefficient of variation of 
$\gamma=0.1$.

To write this up more formal: Given a number of cells (indexed by $i$)
with the *same* pseudotime, we assume that the cells' counts $K_i$ follow 
$$ K_i|Y_i \sim \text{Pois}(s_i Y_i), $$
where the  $Y_i$ are drawn i.i.d. from a distribution with $ \operatorname{E}Y_i=\mu$ and
$\operatorname{Var}K_i=(\gamma\mu)^2$.  Of course, $\operatorname{E}(K_i|Y_i)=\operatorname{Var}(K_i|Y_i)=s_iY_i$.

Using the [law of total expectation](https://en.wikipedia.org/wiki/Law_of_total_expectation),
we find that $\operatorname{E} K_i = s_i \mu$ and from the [law of total variance](https://en.wikipedia.org/wiki/Law_of_total_variance),
that 
$$\begin{align}
\operatorname{Var}K_i &= \operatorname{E}\left(\operatorname{Var}(K_i|Y_i)\right) + \operatorname{Var}\left(\operatorname{E}(K_i|Y_i)\right)\\
&= \operatorname{E}(s_iY_i) + \operatorname{Var}(s_iY_i) \\
&= s_i\operatorname{E}Y_i + s_i^2\operatorname{Var}Y_i\\
&=s_i\mu + s_i^2\gamma^2\mu^2.
\end{align}
$$
For the coefficient of variation, we get
$$ \operatorname{CV}{K_i} = \frac{\sqrt{\operatorname{Var}K_i}}{\operatorname{E}K_i} = \sqrt{ \frac{1}{s_i\mu} + \gamma^2 }$$
We see that the variance of $K_i$ decomposes in two parts: the first captures the Poisson noise and its contribution to the CV drops proportional
to $\sqrt{\mu}$; the second part is constant and described the biological variation (which we deemed to have fixed CV).

In order to get an explicit probability mass function $f_K$ for the $K_i$, we should choose a probability density 
function $f_Y$ for the $Y_i$, so that we can calculate

$$f_K(k) := \text{Prob}(K_i=k ) = \int f_\text{Pois}(k;y)f_Y(y)\text{d}y$$
(where we omitted the size factors $s_i$ for now).

It would seem to be a good choice to use a normal or log-normal pdf for $f_Y$, but, unfortunately, this results in an 
integral that cannot be solved in closed form. However, if we use the [Gamma distribution](https://en.wikipedia.org/wiki/Gamma_distribution)
for $f_Y$, the integra for $f_K$ yields a closed-form expression which is called the *negative-binomial (NB) distribution*, or
the *Gamma-Poisson distribution*. The calculation can be seen [here on Wikipedia](https://en.wikipedia.org/wiki/Negative_binomial_distribution#Gamma–Poisson mixture).

In R, the NB distribution's pmf is given by `dnbinom`, which allows two parametrization. We chose to specify the mean $\mu$ as `mean`
and the parameter `size`, which corresponds to $1/\gamma^2$ in our notation.

We calculate the NB deviance for $\gamma=0.1$ and the corresponding deviance residuals:

```{r}
tbl %>%
mutate( deviance_nb = -2 * ( 
  dnbinom( count, mu = total * pred_expr, size = 1/.1^2, log=TRUE ) - 
  dnbinom( count, mu = count, size = 1/.01, log=TRUE ) ) ) %>%
mutate( devres_nb = sqrt(deviance_nb) * sign( count - total*pred_expr ) ) -> tbl

head(tbl)
```
Now, the gray band is a bit wider:

```{r}
tbl %>% 
ggplot() + 
  geom_point( aes( x=pt, y=y, col=devres_nb ), size=.3 ) +
  geom_line( aes( x=x, y=y ), data=fitted_curve, lty="dotted" ) +
  scale_y_log10() +
  scale_color_gradient2( mid="gray80", limits=c(-2.5,2.5), oob=scales::oob_squish ) +
  theme_bw()
```

Here's the histogram of the NB deviance residuals, and their QQ plot against a standard normal:

```{r}
hist( tbl$devres_nb, 100 )
qqnorm( tbl$devres )
abline( 0, 1 )
```

### Nearest-neighbor smoothing

Smoothing may serve two purposes: (i) to more clearly see the influence of a specific
continuously varying covariate on some measured quantity, or (ii) to remove noise
from a measured quantity.

Now that we know that pseudotime does not explain all of the non-Poisson variance,
we see that our local regression on pseudotime is useful for purpose (i), but to
perform general denoising (ii), we are better with another approach.

A very simple one is this: Let's just average of each cell's $k$ nearest neighbors.

We've got a table of nearest neighbors further ip, using FNN:

```{r}
head( nn$idx )
```

Let's first get vectors of the counts for Aqp4 and for the totals:

```{r}
counts_Aqp4 <- counts[ "Aqp4", ]  
totals <- colSums(counts)
```

Now, let's sum up, for each cell, the Aqp4 counts of its 30 neighbors:

```{r}
counts_Aqp4_nnsum <- unname( rowSums( 
  sapply( 1:ncol(nn$idx), function(j) counts_Aqp4[nn$idx[,j] ] ) ) )
```

Do the same for the totals

```{r}
totals_nnsum <- unname( rowSums( 
  sapply( 1:ncol(nn$idx), function(j) totals[nn$idx[,j] ] ) ) )
```

Fix the cell names:

```{r}
names(counts_Aqp4_nnsum) <- names(counts_Aqp4)
names(totals_nnsum) <- names(totals)
```

Now, calculate the expression fractions as ratio of the neighborhood-summed
Aqp4 counts to the neighborhood-summed totals:

```{r}
tbl %>%
mutate( y_nnsum = ( counts_Aqp4_nnsum / totals_nnsum )[ cell ] ) %>%
mutate( y_nnsum = pmax( y_nnsum, 10^runif( n(), -6.4, -6 ) ) ) -> tbl

head(tbl)
```

Let's first compare this with our fitted curves. The points are now
the neighborhood-smoothed expressions:

```{r}
ggplot(tbl) + 
  geom_point( aes( x=pt, y=y_nnsum ), size=.3, alpha=.4 ) +
  geom_line( aes( x=x, y=y ), data=fitted_curve, col="magenta" ) +
  scale_y_log10() 
```
We see that we now get useful value even in the range below $10^{-4}$.

Here is again the plot of the Aqp4 expression in the UMAP, first using only
the cell's own counts then using the sum over its neighbors:

```{r}
as_tibble(ump) %>%
mutate( y = log( counts_Aqp4/totals * 1e4 + 1 ) ) %>%
ggplot + 
  geom_point( aes( x=V1, y=V2, col=y ), size=.1 ) +
  coord_equal() + 
  scale_color_viridis_c( direction=-1, limits=c(0,3.5), oob=scales::squish )

as_tibble(ump) %>%
mutate( y = log( counts_Aqp4_nnsum/totals_nnsum * 1e4 + 1 ) ) %>%
ggplot + 
  geom_point( aes( x=V1, y=V2, col=y ), size=.1 ) +
  coord_equal() + 
  scale_color_viridis_c( direction=-1, limits=c(0,3.5), oob=scales::squish )
```
We see how the "salt-and-pepper pattern" is smoothed over.

We can also ask whether the smoothing has caused the expression of some cells
to now deviate more strongly than what we would expect if the smoothing only 
corrected for Poisson noise. We can spot such cases of "over-smoothing" by checking
for large Poisson residuals:

```{r}
deviance_nn <- -2 * (
  dpois( counts_Aqp4, totals * counts_Aqp4_nnsum/totals_nnsum, log=TRUE ) -
  dpois( counts_Aqp4, counts_Aqp4, log=TRUE ) )

devres_nn <- sqrt( deviance_nn ) * sign( counts_Aqp4 - totals * counts_Aqp4_nnsum/totals_nnsum )

as_tibble(ump) %>%
mutate( devres = devres_nn ) %>%
ggplot + 
  geom_point( aes( x=V1, y=V2, col=devres ), size=.1 ) +
  coord_equal() + theme_bw() +
  scale_color_gradient2( mid="gray80", limits=c(-4,4), oob=scales::squish ) 
```
### PCA smoothing

Instead of simply averaging over the neighborhood, we can also use
our PCA (calculated at the very beginning of this notebook) to
denoise the data. We simply rotate the PCA scores back into feature
space and undo the centering (as discussed a few weeks ago):

```{r}
expr_pca_smoothed <- t( pca$x %*% t(pca$rotation) ) + pca$center
```

Here's the smoothing result

```{r}
as_tibble(ump) %>%
mutate( y = expr_pca_smoothed["Aqp4",] ) %>%
ggplot + 
  geom_point( aes( x=V1, y=V2, col=y ), size=.1 ) +
  coord_equal() + 
  scale_color_viridis_c( direction=-1, limits=c(0,3.5), oob=scales::squish )
```
To get to the deviance residuals, we have to also undo the log-normalization:

```{r}
pca_pred_counts <- totals * exp( expr_pca_smoothed["Aqp4",] ) / 1e4

deviance_pca <- -2 * (
  dpois( counts_Aqp4, pca_pred_counts, log=TRUE ) -
  dpois( counts_Aqp4, counts_Aqp4, log=TRUE ) )

devres_pca <- sqrt( deviance_pca ) * sign( counts_Aqp4 - pca_pred_counts )

as_tibble(ump) %>%
mutate( devres = devres_pca ) %>%
ggplot + 
  geom_point( aes( x=V1, y=V2, col=devres ), size=.1 ) +
  coord_equal() + theme_bw() +
  scale_color_gradient2( mid="gray80", limits=c(-4,4), oob=scales::squish ) 
```
It's all a bit reddish. Plotting against pseudotime sheds light at the issue:

```{r}
tbl %>%
mutate( y = exp( expr_pca_smoothed["Aqp4",cell] ) / 1e4 ) %>%
ggplot() + 
  geom_point( aes( x=pt, y=y ), size=.3 ) +
  geom_line( aes( x=x, y=y ), data=fitted_curve, col="magenta" ) +
  scale_y_log10() 
```
Because we performed the PCA on log-normalized data where we added the small value $10^{-4}$,
our PCA prediction cannot drop much below this value, and the subtle signal at later pseudotime
gets lost.

#### GLM-PCA

If we replaced the PCA with a procedure that does essentially the same but is able
to work directly with counts, doing away with the needs to mask this, we get a better
fit. 

How to do this will be shown in the nect lecture, but here is already the result:

```{r}
read_lines("Aqp4_glmpca.txt") %>% as.numeric() -> Aqp4_glmpca

head( Aqp4_glmpca )
```
Here's a plot of these "GLM-PCA denoised" values:

```{r}
as_tibble(ump) %>%
  mutate( y = Aqp4_glmpca ) %>%
  ggplot + 
  geom_point( aes( x=V1, y=V2, col=y ), size=.1 ) +
  coord_equal() + 
  scale_color_viridis_c( direction=-1, limits=c(0,3.5), oob=scales::squish )
```

Again, the residuals:

```{r}
glm_pca_pred_counts <- totals * exp( Aqp4_glmpca ) / 1e4

deviance_glm_pca <- -2 * (
  dpois( counts_Aqp4, glm_pca_pred_counts, log=TRUE ) -
  dpois( counts_Aqp4, counts_Aqp4, log=TRUE ) )

devres_glm_pca <- sqrt( deviance_glm_pca ) * sign( counts_Aqp4 - glm_pca_pred_counts )

as_tibble(ump) %>%
mutate( devres = devres_glm_pca ) %>%
ggplot + 
  geom_point( aes( x=V1, y=V2, col=devres ), size=.1 ) +
  coord_equal() + theme_bw() +
  scale_color_gradient2( mid="gray80", limits=c(-4,4), oob=scales::squish ) 
```
The plot against the pseudotime now dips down deeper, and follows the fit well
much longer:

```{r}
tbl %>%
mutate( y = exp( Aqp4_glmpca[cells$celltype1=="lineage"] ) / 1e4 ) %>%
ggplot() + 
  geom_point( aes( x=pt, y=y ), size=.3, alpha=.3 ) +
  geom_line( aes( x=x, y=y ), data=fitted_curve, col="magenta" ) +
  scale_y_log10() 
```
What's happening at the end? Is this the fault of locfit or of GLM-PCA. This 
is not clear at this point, so we leave the question open for now.

### Autoencoder

The next step over the GLM-PCA will be to replace it with a non-linear procedure.
In one of the next lectures, we will learn how such an autoencoder works, but here is already
the result.

Load the autoencoder-denoised data:

```{r}
read_lines("Aqp4_ae.txt") %>% as.numeric() -> Aqp4_ae_pred_counts
```

Plot it in a UMAP:

```{r}
as_tibble(ump) %>%
  mutate( y = log(Aqp4_ae_pred_counts/totals*1e4) ) %>%
  ggplot + 
  geom_point( aes( x=V1, y=V2, col=y ), size=.1 ) +
  coord_equal() + 
  scale_color_viridis_c( direction=-1, limits=c(0,3.5), oob=scales::squish )
```
The residuals:

```{r}
deviance_ae <- -2 * (
  dpois( counts_Aqp4, Aqp4_ae_pred_counts, log=TRUE ) -
  dpois( counts_Aqp4, counts_Aqp4, log=TRUE ) )

devres_ae <- sqrt( deviance_ae ) * sign( counts_Aqp4 - Aqp4_ae_pred_counts )

as_tibble(ump) %>%
mutate( devres = devres_ae ) %>%
ggplot + 
  geom_point( aes( x=V1, y=V2, col=devres ), size=.1 ) +
  coord_equal() + theme_bw() +
  scale_color_gradient2( mid="gray80", limits=c(-4,4), oob=scales::squish ) 
```

And the plot against pseudotime

```{r}
tbl %>%
  mutate( y = Aqp4_ae_pred_counts[cells$celltype1=="lineage"] / total * 1e4 ) %>%
  ggplot() + 
  geom_point( aes( x=pt, y=y ), size=.3, alpha=.3 ) +
  geom_line( aes( x=x, y=y*1e4 ), data=fitted_curve, col="magenta" ) +
  scale_y_log10() 
```
It seems our autoencoder has a floor, too, although lower than before. We fix this when we revisit it
in a later lecture.
