---
title: "Single-cell analysis II"
format: html
---

## Course "Advanced Computational Systems Biology"

University of Heidelberg, summer term 2023

## Single-cell RNA-Seq data analysis
### Day 2: Pseudotime

Simon Anders

### Example data

This time, we use data from this paper:

D Carvajal Ibañez, M Skabkin, J Hooli et al.: *Interferon regulates neural stem cell function at all ages by orchestrating mTOR and cell cycle*. EMBO Molecular Medicine (2023) 15:e16434. https://doi.org/10.15252/emmm.202216434

The data is here: https://papagei.bioquant.uni-heidelberg.de/simon/Vl23sysbiol/ifnagrko/

We load the data as before:

```{r}
suppressPackageStartupMessages({
  library( Matrix )
  library( Seurat )
  library( tidyverse ) })

mtx_mm <- read.delim( gzfile( "/mnt/raid/ifnagrko/ifnagrko_raw_counts.mtx.gz" ), 
    comment.char="%", sep=" ", header=FALSE )

cells <- read.csv( "/mnt/raid/ifnagrko/ifnagrko_obs.csv", row.names = 1 )
genes <- read.csv( "/mnt/raid/ifnagrko/ifnagrko_var.csv", row.names = 1 )
counts <- readMM( "/mnt/raid/ifnagrko/ifnagrko_raw_counts.mtx.gz" )
rownames(counts) <- rownames(cells)
colnames(counts) <- genes$gene_name
```

### Standard analysis

Run a standard Seurat analysis, as on previous day

```{r}
set.seed( 12345678 )

seu <- CreateSeuratObject( 
  t(counts), 
  meta.data = cells )

seu <- NormalizeData( seu )
seu <- FindVariableFeatures( seu )
seu <- ScaleData( seu )
seu <- RunPCA( seu )
seu <- RunUMAP( seu, dims=1:20 )

UMAPPlot( seu, group.by="celltype", label=TRUE) + coord_equal()
```


### Looking at pseudotime in the "lineage"

For the cells within the "lineage", a pseudotime has been calculated. Plot it:

```{r}
bind_cols(
  seu@reductions$umap@cell.embeddings,
  dpt = seu@meta.data$dpt_pseudotime ) %>%
ggplot +
  geom_point( aes( x=UMAP_1, y=UMAP_2, col=dpt ), size=.1 ) +
  coord_equal() +
  scale_color_viridis_c()
```

Our UMAP looks a bit different from the one the paper's authors have calculated
and placed in their meta data table:

```{r}
seu@meta.data %>%
  ggplot + geom_point( aes( x=UMAP1, y=UMAP2, col=dpt_pseudotime ), size=.1) +
  coord_equal() + scale_color_viridis_c()
```

The dividing cells (marked by Ki-67) are here:

```{r}
FeaturePlot( seu, "Mki67" )
```

### Regression cell expression on the lineage

The astrocyte marker Aqp4 is high in early pseudotime and goes down quickly

```{r}
FeaturePlot( seu, "Aqp4" )
```
Plot its expression agains pseudotime

```{r}
tibble(
   expr = seu@assays$RNA@data["Aqp4",],
   pt = seu@meta.data$dpt_pseudotime ) %>%
mutate( expr = ifelse( expr > 0, expr, runif( n(), -.2, 0 ) ) )  %>%
ggplot +
  geom_point( aes( x=pt, y=expr ), size=.1, alpha=.3 )
```

Here, we have expanded the cells with zero expression to a narrow band, to
reduce overplotting

### Averages per bin

We divide the pseudotime into 20 "bins" and average the expression of Aqp4
over the cells of each bin. We do this for each of the 4 samples separately

```{r}
tibble(
  expr = seu@assays$RNA@data["Aqp4",],
  pt = seu@meta.data$dpt_pseudotime,
  genotype = seu@meta.data$genotype,
  age = seu@meta.data$age ) %>%
mutate( ptbin = round( pt*20 )/20 ) %>%
group_by( ptbin, genotype, age  ) %>%
summarise( mean(expr), n() ) %>%
ggplot +
  geom_line( aes( x=ptbin, y=`mean(expr)`, 
      col=genotype, linetype=age ) )
```

For comparison, here is the expression as feature plot, split by sample:

```{r}
bind_cols(
   seu@reductions$umap@cell.embeddings,
   expr = seu@assays$RNA@data["Aqp4",],
   seu@meta.data ) %>% 
mutate( ptbin = round(dpt_pseudotime*30)/30 ) %>%
ggplot + 
  geom_point( aes( x=UMAP_1, y=UMAP_2, col=expr ), size=.1 ) +
  coord_equal() + scale_color_viridis_c(direction=-1) +
  facet_grid( age ~ genotype )
```

### Local regression

A smoother curve can be found by performing a running mean (i.e.,
averaging over a sliding window), by performin kernel smoothing
(weighting the cells within the sliding window according to their
distance to the window center), or by performing local regression
(also known as LOWESS or LOESS smoothing).

To keep things simple, we will not split by sample this time.

Here's the table of the data we'll need.

```{r}
tibble(
  expr = seu@assays$RNA@data["Aqp4",],
  count = seu@assays$RNA@counts["Aqp4",],
  total = colSums( seu@assays$RNA@counts ),
  pt = seu@meta.data$dpt_pseudotime ) %>%
mutate( expr = log( count/total * 1e4 + 1 ) ) %>%
filter( !is.na(pt) ) -> tbl

head( tbl )
```

Here's the LOESS fit:

```{r}
library(locfit)

fit <- locfit( expr ~ lp(pt,nn=.1), tbl, 
          ev=lfgrid( 100, 0, 1 ) )  

plot( tbl$pt, tbl$expr, col="gray")
lines( fit$eva$xev, predict(fit) )
```
Explanation: The locfit function regresses the log-normalized expression (column expr)
onto the pseudotime (column pt), hence `expr ~ pt`, using the data in `tbl`. For each
x value of the curve, the y value is calculated by taking all data points under the kernel,
weight them according to the kernel, fit a regression line, and keep only the one point of the regression
line at the x position we are caclulating. The kernel's width is adjusted such that
it always covers 10% of the data points (`nn=0.1`).

This is calculated for 100 points between 0 and 1 (`lfgrid` argument), and these 100
equi-spaced values are then in `fit$eva$xev`. the corresponding y coordinates are in 
`predict(fit)`.

### Local GLM regression

Our expression values are distorted by the pseudocount in the log-normalization. The
scaling factor (`1e4`, see previous day) directly influences the gap between the zeroes
and the other values, which is artifactual.

Making locfit aware that this is Poisson-distributed data and providing it with the
raw data (`count` and `total` instead of `expr`) rectifies all this:

```{r}
fit <- locfit( count ~ lp(pt,nn=.1), tbl, 
               family="poisson", base=log(tbl$total),
               ev=lfgrid( 100, 0, 1 ) )  
plot( tbl$pt, tbl$expr, col="gray", ylim=c(-4,4) )
lines( seq(0,1,l=100), log( predict(fit) * 1e4 ) )
```

Now, can see that expression keeps falling, with cells with non-zero expression getting more and more rare.

### Several genes

Here is the code from the previous section in a function:

```{r}
locfitpois <- function( counts, totals, pseudotimes ) {
  fit <- locfit( 
    counts ~ lp( pseudotimes, nn=.1 ),  
        ev = lfgrid( 100, 0, 1 ),
        family = "poisson",
        base = log(totals) )
  tibble( 
     x = fit$eva$xev,
     y = predict(fit) )
}
```

Testing it:

```{r}
gene <- "Lhx2"

fit <- locfitpois( 
  seu@assays$RNA@counts[gene,], 
  seu@meta.data$nCount, 
  seu@meta.data$dpt_pseudotime )

plot(   
   seu@meta.data$dpt_pseudotime,
   seu@assays$RNA@data[gene,], 
   col = "gray", pch=".", ylim=c(-4,4) )

lines( fit$x, log(fit$y*1e4) )
```

Let's run this for several genes:

```{r}
genes <- c( "Aqp4", "Mki67", "Lhx2")

genes %>% set_names() %>%
map_dfc( function(gene)
  locfitpois( 
    seu@assays$RNA@counts[gene,], 
    seu@meta.data$nCount, 
    seu@meta.data$dpt_pseudotime )$y ) -> m

head( m )
```

Image of this matrix

```{r}
image(as.matrix(m))
```
With proper labels:

```{r}
image( as.matrix(m), xlab="pseudotime", yaxt="n" )
axis( 2, seq( 0, 1, length.out=ncol(m) ), colnames(m), las=2 )
```
We should make such a heatmap with a few more interesting genes to learn something about 
how the cells' expression patterns change during the path through pseudotime.

### Further reading

- A lecture on kernel smoothing and local regression: https://papagei.bioquant.uni-heidelberg.de/simon/Vl23sco/smoothing.html

- A lecture on diffusion pseudotime: https://papagei.bioquant.uni-heidelberg.de/simon/Vl23sco/diffdist.html
