---
title: "Single-cell analysis I"
format: html
---

## Course "Advanced Computational Systems Biology"

University of Heidelberg, summer term 2023

## Single-cell RNA-Seq data analysis
### Day 1: Standard analysis with Seurat

Simon Anders

### Note

In this document I have put together the R code that we have discussed in the 
two days of the course that I have tought. This is just meant as a reminder, not as a
self-contained tutorial

### Example dataset

As example data, we have used the RNA-Seq part of the benchmark dataset produced
for this paper: 

Luecken et al. (2022): *Benchmarking atlas-level data integration in single-cell genomics*.
Nature Methods, 19:41. https://doi.org/10.1038/s41592-021-01336-8

We use the Seurat package (https://satijalab.org/seurat/) for data analysis, and
the matrix package to load the data:

```{r}
suppressPackageStartupMessages({
  library(Matrix)
  library(ggplot2)
  library(Seurat)
})
```

### Loading the data

The data is available here: https://papagei.bioquant.uni-heidelberg.de/simon/Vl23sysbiol/Luecken2021_GEx/

The big files contain all cells, the cells with `s4d1` only those from sample 4 / donor 1. We load them:

```{r}
obs <- read.csv( "/mnt/raid/anders/tmp/neurips2021BMMC_10XMultiome_GEX/obs_s4d1.csv.gz", row.names=1 )
var <- read.csv( "/mnt/raid/anders/tmp/neurips2021BMMC_10XMultiome_GEX/var.csv.gz", row.names=1 )
counts <- readMM( "/mnt/raid/anders/tmp/neurips2021BMMC_10XMultiome_GEX/counts_s4d1.mtx.gz" )
```

Here's the data

The count matrix (top-left corner; only zeroes), and the matrix dimensions:

```{r}
counts[ 1:5, 1:5 ]
cat( "\n" )

dim( counts )
```
The cell annotation

```{r}
dim( obs )

head(obs)
```

The gene annotation

```{r}
dim( var )

head(var)
```

Copy over cell and gene names:

```{r}
rownames(counts) <- rownames(obs)
colnames(counts) <- rownames(var)
```

Create a Seurat object:

```{r}
seu <- CreateSeuratObject( 
  t(counts), 
  meta.data = obs[,c("cell_type","batch","Samplename","DonorNumber") ] )
```

### Standard Seurat analysis

Run a standard analysis

```{r}
set.seed( 12345678 )

seu <- NormalizeData( seu )
seu <- FindVariableFeatures( seu )
seu <- ScaleData( seu )
seu <- RunPCA( seu )
seu <- RunUMAP( seu, dims=1:20 )
seu <- FindNeighbors( seu, dims = 1:20 )
seu <- FindClusters( seu, resolution = 0.5 )
```

The UMAP:

```{r}
UMAPPlot( seu, label=TRUE ) + coord_equal()
```

### Sleepwalk

Inspect the UMAP with Sleepwalk (not run):

```r
sleepwalk::sleepwalk( 
  seu@reductions$umap@cell.embeddings,
  seu@reductions$pca@cell.embeddings )
```

Here, we have pulled from the Seurat object the 2D UMAP coordinates for all cells
and the 20-dim PCA coordinates.

For more on Sleepwalk, see here: https://anders-biostat.github.io/sleepwalk/


### Inspection 

Next, we inspect the object and make a few comparisons.

First: For each cell, the total count (sum over all genes) is in the meta-data column
`nCount_RNA`. Compare that this is the same as what we get if we add up the counts
ourselves. 

```{r}
plot( colSums( seu@assays$RNA@counts ), seu@meta.data$nCount_RNA )
```
Next, we compare the normalized data (in `seu@assays$RNA@data`) with computing the 
normalization ourselves, using the formula $y=\log(k/s \cdot 10^4 + 1)$, with $k$
the count for a gene, $s$ the total count of the cell (summed over all the genes),
and $y$ the log-normalized data. To see that this is infact the formula that Seurat has used
we compare (for the gene ACTB):

```{r}
plot(
  log( seu@assays$RNA@counts["ACTB",] / seu@meta.data$nCount_RNA * 1e4 + 1 ),
  seu@assays$RNA@data["ACTB",] )
```
A histogram of th expression of ACTB:

```{r}
hist( seu@assays$RNA@data["ACTB",] )
```

### Gene expression means by cluster

The clusters are stored here:

```{r}
head( seu@meta.data$seurat_clusters )
```

Task: What is the average expression (on the log scale) of ACTB in the cells of Cluster 3?

Solution: 

```{r}
# Which cells are in cluster 3
w <- seu@meta.data$seurat_clusters == 3
head( w )
```

```{r}
# Get the row for ACTB and all columns from cluster 3
head( seu@assays$RNA@data[ "ACTB", w ]  )
```

```{r}
# Take the mean of all these values
mean( seu@assays$RNA@data[ "ACTB", w ] )
```
How to get the means for cluster 3 for all genes?

```{r}
# Take all rows (all genes),  but only the columns for cells in w
seu@assays$RNA@data[ , w ][ 1:5, 1:5 ]
cat( "\n\n\n" )

# Take the mean of each row
head( rowMeans( seu@assays$RNA@data[ , w ] ) )
```
Here is a function that does all this:

```{r}
# A function to get gene means for a cluster
gene_means_for_cluster <- function(cl) {
  w <- seu@meta.data$seurat_clusters == cl
  rowMeans( seu@assays$RNA@data[ , w ] )
}
```

Try it out, look at ACTB

```{r}
gene_means_for_cluster(3)["ACTB"]
```

Call this function for all clusters

```{r}
#Call this function for all clusters
gene_means <- sapply( 0:14, gene_means_for_cluster )

# Set column names
colnames(gene_means) <- sprintf( "Cl_%02d", 0:14 )
gene_means[1:5,1:5]
```

Indexing this matrix:

```{r}
gene_means["ACTB",3]
gene_means["ACTB","Cl_03"]
```

### Finding interesting genes

Which genes vary a lot across clusters?

```{r}
library( matrixStats )

gene_vars_across_clusters <- rowVars( gene_means, useNames=TRUE )

head( gene_vars_across_clusters )
```
Sort this:

```{r}
sort( gene_vars_across_clusters, decreasing=TRUE )[1:20]
```
Store the top 50:
```{r}
interesting_genes <- names( sort( 
  gene_vars_across_clusters, decreasing=TRUE )[1:50] )
```

Center the gene means:

```{r}
centered_gene_means <- gene_means - rowMeans(gene_means)
```

Make a heatmap

```{r}
heatmap( 
  centered_gene_means[ interesting_genes, ], 
  scale="none", zlim=c(-4,4),
  col=colorRampPalette(c("blue","gray","red"))(100) )
```

### Identifying cell types

From this heatmap, we can identify the cell types in the clusters. Clusters 6 and 14
clearly contain erythrocytes and their progenitors. (They express hemoglobin genes.)

We can also check for known marker genes

```{r}
lymphoid_genes <- c( "CD3D", "CD3E", "CD3G", "CD19", "MS4A1", "PAX5", "TRA", "TRB", "TRG", "TRD", "CD79A", "CD79B", "RAG1", "RAG2" )  # provided by ChatGPT

genes <- intersect( lymphoid_genes, rownames(seu) )

heatmap( 
  centered_gene_means[ genes, ], 
  scale="none", zlim=c(-4,4),
  col=colorRampPalette(c("blue","gray","red"))(100) )
```
Cluster 1 and 4 are T cells

The data already has a cell-type annotation. We can compare the overlap between
the cell-type annotation (rows below) from the paper authors with our Seurat cluster (columns):

```{r}
table(
   seu@meta.data$cell_type,
   seu@meta.data$seurat_clusters ) 
```

Let's go on with the authors' cell-type assignment.

```{r}
UMAPPlot( seu, group.by="cell_type", label=TRUE ) + coord_equal()
```

#### Expression differences between cell types

Which cells are CD4+ T cells and wich are CD8+ cell types?

```{r}
wCD8 <- seu@meta.data$cell_type == "CD8+ T"
wCD4 <- substr( seu@meta.data$cell_type, 1, 6 ) == "CD4+ T"
```

Calculate for all genes their mean expression in these two cell types:
```{r}
means_CD8 <- rowMeans( seu@assays$RNA@data[ , wCD8 ] )
means_CD4 <- rowMeans( seu@assays$RNA@data[ , wCD4 ] )
```

Compare them:

```{r}
plot( means_CD4, means_CD8, asp=1 )
abline(0,1)
```

We can also rotate the plot by 45° to get an MA plot:

```{r}
plot( 
  ( means_CD4 + means_CD8 ) /2, 
  means_CD8 - means_CD4, cex=.3 )
```
What might be the gene at the top?

```{r}
head( sort( means_CD8 - means_CD4, decreasing=TRUE ) )
```
If we compare this gene's expression between the two groups by t-test

```{r}
t.test(
  seu@assays$RNA@data[ "AOAH", wCD8 ],
  seu@assays$RNA@data[ "AOAH", wCD4 ] )
```
The low p value is, however, rather meaningless: As these were two different cell clusters 
in the *same* sample, it is expected to see strong differences.

### Testing across multiple samples

However, we have multiple samples.

It makes more sense to calculate the mean expression of AOAH in the CD4+ and CD8+ t cells
from each sample:

```{r}
# Load the *full* data set:
  
obs <- read.csv( "/mnt/raid/anders/tmp/neurips2021BMMC_10XMultiome_GEX/obs.csv.gz", row.names=1 )
var <- read.csv( "/mnt/raid/anders/tmp/neurips2021BMMC_10XMultiome_GEX/var.csv.gz", row.names=1 )
counts <- readMM( "/mnt/raid/anders/tmp/neurips2021BMMC_10XMultiome_GEX/counts.mtx.gz" )

rownames(counts) <- rownames(obs)
colnames(counts) <- rownames(var)
```

Make a table with the needed data, for the gene AOAH, with one row per cell:

```{r}
suppressPackageStartupMessages( 
  library( tidyverse ) )

tibble(
  cell = rownames(counts),
  gene_count = counts[,"AOAH"],
  total = rowSums(counts),
  cell_type = obs$cell_type,
  batch = obs$batch,
  donor = obs$DonorNumber ) -> tbl

head( tbl )
```  

Now, we can subset only to the T cells, then average over the cells from each donor

```{r}
tbl %>%
filter( cell_type %in% c( "CD4+ T naive", "CD4+ T activated", "CD8+ T" ) ) %>%
mutate( simple_cell_type = str_sub( cell_type, 1, 6 ) ) %>%
mutate( expr = log( gene_count / total * 1e4 + 1 ) ) %>%
group_by( simple_cell_type, donor ) %>%
summarise( mean(expr) ) %>%
pivot_wider( names_from = "simple_cell_type", values_from = "mean(expr)" ) -> tbl2

tbl2
```
We compare these with aa paired t test

```{r}
t.test( tbl2$`CD4+ T`, tbl2$`CD8+ T`, paired=TRUE )
```
Now, we have a actually useful p value, rejecting the null hypothesis that there
is no difference in the expression of AOAH between CD4+ and CD8+ t cells.

Here, we have first log-normalized the data, then averaged over cells.

There are some arguments in favour of summing up the counts over the cells and dividing 
by the sum of the totals instead, and taking the log of that. This is called making a "pseudo-bulk":

```{r}
tbl %>%
filter( cell_type %in% c( "CD4+ T naive", "CD4+ T activated", "CD8+ T" ) ) %>%
mutate( simple_cell_type = str_sub( cell_type, 1, 6 ) ) %>%
group_by( simple_cell_type, donor ) %>%
summarise( m = log( sum(gene_count)/sum(total) ) ) %>%
pivot_wider( names_from = "simple_cell_type", values_from = "m" ) -> tbl2b
  
t.test( tbl2b$`CD4+ T`, tbl2b$`CD8+ T`, paired=TRUE )
```
